---
title: "Objective 4"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/chanamessinger/Dropbox/Columbia/Practicum/Data/csv")
getwd()

packages <- c("tmap", "devtools","knitr","tidyverse","broom", "DT","ggmap", "ggthemes", "maps", "statebins", "stringr","scales","tidyverse","readxl","statebins","RJSONIO","XML","RCurl", "readr", "tidyr", "dplyr", "ggplot2", "rgdal", "leaflet", "RColorBrewer", "rqog", "censusr", "rtweet")


packages <- lapply(packages, FUN = function(x) {
  if(!require(x, character.only = TRUE)) {
    install.packages(x)
  library(x, character.only = TRUE)
  }
}
)
library(zipcode)
```

Extract zipcode from patient adddress, then inner join it on the meds and extract latitude and longitude
```{r}
setwd("/Users/chanamessinger/Dropbox/Columbia/Practicum/spring2019/My Data")
# save_as_csv(meds_bc, file_name = "meds_bc")
# save_as_csv(patient_bc, file_name = "patient_bc")

meds_bc <- read_csv("meds_bc.csv")
patient_bc <- read_csv("patient_bc.csv")


meds_bc <- meds_bc %>%
  rename(ID = PATIENT)

patient_bc_info <- patient_bc %>%
  select("ID", "ADDRESS") %>%
  mutate(zip = as.character(str_sub(str_extract_all(ADDRESS, "[A-Z]{2} [0-9]{5}"), 4)))
  

data(zipcode)
zipcode$zip <- as.character(zipcode$zip)

patient_bc_info <- inner_join(patient_bc_info, zipcode, by = "zip")

patient_bc_info <- inner_join(meds_bc, patient_bc_info, "ID") %>%
  select(-"REASONCODE", -"REASONDESCRIPTION")
```

Create a factor for the type of BC
```{r}
patient_bc_info <- patient_bc_info %>%
  mutate(
    BCTYPE = case_when(
    CODE.bc %in% c(1000158, 1000128) ~ "Injection",
    CODE.bc %in% c(748962, 831533, 749785, 749762, 751905, 748856, 1359133, 757594, 748879, 235389) ~ "Pills",
    CODE.bc %in% c(1605257, 1856546, 807283, 1111011, 646250) ~ "LARC",
    CODE.bc == 1367439 ~ "Ring",
    TRUE ~ "Other"
    )
  )

patient_bc_info$BCTYPE <- as.factor(patient_bc_info$BCTYPE)
```


Put a population column for each geoid (county/tract)
```{r}
setwd("/Users/chanamessinger/Dropbox/Columbia/Practicum/spring2019/My data")
population <- read_csv("censustract.csv")
population <- population %>%
  select("GEOID", "POP10") %>%
  rename("Population" = "POP10", "TRACTCE" = "GEOID") %>%
  mutate(TRACTCE = as.numeric(TRACTCE))
```

Append census tract population to info dataframe.
```{r}
patient_bc_info <- patient_bc_info %>%
  rename(lat = latitude, lon = longitude)

patient_bc_info <- left_join(patient_bc_info, population, by = "TRACTCE") %>%
  rename("TractPop" = "Population")



# patient_bc_info <- append_geoid(patient_bc_info, "tr")
# patient_bc_info <- patient_bc_info %>%
#   rename(TRACTCE = geoid)
# patient_bc_info <- append_geoid(patient_bc_info, "co")
# could have taken above from the tract

# patient_bc_info <- patient_bc_info %>%
#   rename(COUNTYFP = geoid)
# setwd("/Users/chanamessinger/Dropbox/Columbia/Practicum/spring2019/My data")
# save_as_csv(patient_bc_info, file_name = "patient_bc_info")
```
Start here

```{r}
setwd("/Users/chanamessinger/Dropbox/Columbia/Practicum/spring2019/My data")
patient_bc_info <- read_csv("patient_bc_info.csv")
```


Get most common birth control by county
```{r}
bc_by_county <- patient_bc_info %>%
  group_by(`COUNTYFP`) %>%
  count(BCTYPE) %>%
  spread(BCTYPE, n) %>%
  mutate(PILLPROP = Pills/sum(Pills, Ring, Injection, LARC, na.rm = T)) %>%
  ungroup() %>%
  mutate(COUNTYFP = str_sub(COUNTYFP, 3))
```

Get most common birth control by census tracts
```{r}
bc_by_tract <- patient_bc_info %>%
  group_by(`TRACTCE`) %>%
  count(BCTYPE) %>%
  spread(BCTYPE, n) %>%
  mutate(PILLPROP = Pills/sum(Pills, Ring, Injection, LARC, na.rm = T)) %>%
  ungroup() %>%
  mutate(TRACTCE = str_sub(TRACTCE, 6))

bc_by_tract2 <- distinct(bc_by_tract, TRACTCE, .keep_all = TRUE)

```

Map Function

```{r}
create_map <- function(level = "co", data, fill){
  setwd("/Users/chanamessinger/Dropbox/Columbia/Practicum/spring2019/My data/Shape Files")
  mass_census <- readOGR("mass_census/.","cb_2017_25_tract_500k")
  
   if (level %in% c("county", "co", "County")) {
    gradation = "COUNTYFP" 
    } else if (level %in% c("tract", "tr", "Tract", "Census Tract", "censustract", "census tract")) {
      gradation = "TRACTCE" 
      } else {
        print("Try co or tr for level")
      }
  
  shape_with_data <- mass_census@data %>% 
  left_join(data, by = gradation)
  
  mass_census@data <- shape_with_data
  
  tm <- tm_shape(mass_census) + tm_borders() + tm_fill(fill)
  tm
  return(tm)
}
```



Map with county shape files (not working)
```{r}
# setwd("/Users/chanamessinger/Dropbox/Columbia/Practicum/spring2019/My data/Shape Files")
# county_map <- readOGR("tl_2018_us_county/.","tl_2018_us_county")
# county_map@data <- filter(county_map@data, STATEFP == 25)
# 
# shape_with_data <- county_map@data %>% 
#   left_join(bc_by_county, by = "COUNTYFP")
# 
# county_map@data <- shape_with_data
# 
# tm_shape(county_map) + tm_borders() + tm_fill()
# 
# 
# county_map2 <- merge(county_map, bc_by_county, by="COUNTYFP")

```

Maps
```{r}
create_map("co", bc_by_county, "PILLPROP")
create_map("tr", bc_by_tract2, "PILLPROP")
```

```{r}
bc_by_county2 <- patient_bc_info %>%
  group_by(`COUNTYFP`) %>%
  count(BCTYPE) %>%
  spread(BCTYPE, n) %>%
  mutate(LARCPROP = LARC/sum(Pills, Ring, Injection, LARC, na.rm = T)) %>%
  ungroup() %>%
  mutate(COUNTYFP = str_sub(COUNTYFP, 3))


bc_by_tract2 <- patient_bc_info %>%
  group_by(`TRACTCE`) %>%
  count(BCTYPE) %>%
  spread(BCTYPE, n) %>%
  mutate(LARCPROP = LARC/sum(Pills, Ring, Injection, LARC, na.rm = T)) %>%
  ungroup() %>%
  mutate(TRACTCE = str_sub(TRACTCE, 6)) %>%
  distinct(TRACTCE, .keep_all = TRUE)

```

```{r}
create_map("co", bc_by_county2, "LARCPROP")
create_map("tr", bc_by_tract2, "LARCPROP")

```
(Second most common contraceptive method in each county)

```{r}
bc_by_county3 <- patient_bc_info %>%
  group_by(`COUNTYFP`) %>%
  count(BCTYPE) %>%
  spread(BCTYPE, n) %>% 
  ungroup() %>%
  select(-COUNTYFP) %>%
  rownames_to_column() %>%
  gather(column, value, -rowname) %>%
  group_by(rowname) %>% 
  mutate(rk = rank(-value, ties.method = "first")) %>%
  filter(rk == 2) %>% 
  arrange(as.numeric(rowname), rk) 

bc_by_county4 <- patient_bc_info %>%
  group_by(`COUNTYFP`) %>%
  count(BCTYPE) %>%
  spread(BCTYPE, n) %>% 
  ungroup() %>%
  rownames_to_column() %>%
  select(rowname, COUNTYFP)

bc_by_county3 <- inner_join(bc_by_county3, bc_by_county4, by = "rowname") %>%
  ungroup() %>%
  select(COUNTYFP, column) %>%
  rename("SecondMost" = "column") %>%
  mutate(SecondMost = as.factor(SecondMost)) %>%
  mutate(COUNTYFP = str_sub(COUNTYFP, 3))
```

```{r}
bc_by_tract3 <- patient_bc_info %>%
  group_by(`TRACTCE`) %>%
  count(BCTYPE) %>%
  spread(BCTYPE, n) %>% 
  ungroup() %>%
  select(-TRACTCE) %>%
  rownames_to_column() %>%
  gather(column, value, -rowname) %>%
  group_by(rowname) %>% 
  mutate(rk = rank(-value, ties.method = "first")) %>%
  filter(rk == 2) %>% 
  arrange(as.numeric(rowname), rk) 

bc_by_tract4 <- patient_bc_info %>%
  group_by(`TRACTCE`) %>%
  count(BCTYPE) %>%
  spread(BCTYPE, n) %>% 
  ungroup() %>%
  rownames_to_column() %>%
  select(rowname, TRACTCE)

bc_by_tract3 <- inner_join(bc_by_tract3, bc_by_tract4, by = "rowname") %>%
  ungroup() %>%
  select(TRACTCE, column) %>%
  rename("SecondMost" = "column") %>%
  mutate(SecondMost = as.factor(SecondMost)) %>%
  mutate(TRACTCE = str_sub(TRACTCE, 6)) %>%
  distinct(TRACTCE, .keep_all = TRUE)
  
```



```{r}
create_map("co", bc_by_county3, "SecondMost")
create_map("tr", bc_by_tract3, "SecondMost")

```
Data cleaning
```{r}
# setwd("/Users/chanamessinger/Dropbox/Columbia/Practicum/spring2019/My data")
# 
# census_data_healthcare <- read_csv("ECN_2012_US_62Z1_with_ann.csv")
# census_data_healthcare_sub <- census_data_healthcare[-1,]
# 
# offices <- c("Health care and social assistance", "HMO medical centers", "Hospitals", "Offices of physicians")
# 
# census_data_healthcare_sub <- census_data_healthcare %>%
#   rename("zip" = "GEO.id2", "Office Type" = "NAICS.display-label") %>%
#   select(zip, `Office Type`, ESTAB) %>%
#   mutate(ESTAB = as.numeric(ESTAB)) %>%
#   group_by(zip, `Office Type`) %>%
#   summarise(ESTABS = sum(ESTAB)) %>%
#   ungroup() %>%
#   filter(`Office Type` %in% offices)
# 
# census_estabs <- census_data_healthcare_sub %>%
#   group_by(zip) %>%
#   summarize(Total = sum(ESTABS))
# 
# census_estabs <- inner_join(census_estabs, zipcode, by = "zip") %>%
#   rename("lat" = "latitude", "lon" = "longitude")
# 
# census_estabs <- append_geoid(census_estabs, "tr")
# census_estabs <- census_estabs %>%
#   mutate(COUNTYFP = str_sub(geoid, 1, 5)) %>%
#   rename(TRACTCE = "geoid")
# 
# setwd("/Users/chanamessinger/Dropbox/Columbia/Practicum/spring2019/My data")
# save_as_csv(census_estabs, file_name = "census_estabs")
```


```{r}
setwd("/Users/chanamessinger/Dropbox/Columbia/Practicum/spring2019/My data")
census_estabs <- read_csv("census_estabs.csv")


census_estabs2 <- census_estabs %>%
  select(COUNTYFP, Total) %>%
  group_by(COUNTYFP) %>%
  summarize(`Total Health Care Establishments` = sum(Total)) %>%
  mutate(COUNTYFP = str_sub(COUNTYFP, 3))

patient_bc_info2 <- select(patient_bc_info, TRACTCE, TractPop)  %>%
  mutate(TRACTCE = as.character(TRACTCE))

census_estabs2_tract <- census_estabs %>%
  select(TRACTCE, Total) %>%
  inner_join(patient_bc_info2, by = "TRACTCE") %>%
  group_by(TRACTCE, TractPop) %>%
  summarize(`Total Health Care Establishments` = sum(Total)) %>%
  ungroup() %>%
  mutate(`Health Care Establishments per Capita` = `Total Health Care Establishments`/TractPop) %>%
  mutate(TRACTCE = str_sub(TRACTCE, 6)) %>%
  distinct(TRACTCE, .keep_all = TRUE)
```

```{r}
create_map("co", census_estabs2, "Total Health Care Establishments")
create_map("tr", census_estabs2_tract, "Health Care Establishments per Capita")

```
(health care per capita)
```{r}
patient_sub <- patients %>%
  select("ID", "ADDRESS")

patient_sub$zip <- str_extract_all(patient_sub$ADDRESS, "[A-Z]{2} [0-9]{5}")
patient_sub$zip <- str_sub(patient_sub$zip, 4)
patient_sub$zip <- as.character(patient_sub$zip)

patient_sub <- inner_join(patient_sub, zipcode, by = "zip")


patient_sub <- patient_sub %>%
  rename(lat = latitude, lon = longitude)

#patient_census <- append_geoid(patient_bc_type, "co")

patient_census <- read_csv("/Users/chanamessinger/Dropbox/Columbia/Practicum/spring2019/My data/patient_census.csv")

patients <- read_csv("/Users/chanamessinger/Dropbox/Columbia/Practicum/Data/csv/patients.csv")

patient_census <- patient_census %>%
  select(ID, lat, lon, COUNTYFP)
save_as_csv(patient_census, file_name = "patient_census")

patient_census <- patient_bc_info %>%
  select(ID, lat, lon, COUNTYFP, TRACTCE)
```


```{r}
observations <- read_csv(file.choose())

obs_sub <- observations %>%
  filter(DESCRIPTION == "Body Mass Index") %>%
  select(-CODE, -UNITS, -DESCRIPTION) %>%
  group_by(PATIENT) %>%
  filter(DATE == min(as.Date(DATE))) %>%
  rename(ID = PATIENT)


obs_sub <- left_join(obs_sub, patient_census, by = "ID") %>%
  filter(is.na(COUNTYFP) == F) %>%
  distinct()

bmi <- obs_sub %>%
  mutate(BMI = as.numeric(VALUE)) %>%
  group_by(COUNTYFP) %>%
  summarize(mean_bmi = mean(BMI)) %>%
  mutate(COUNTYFP = str_sub(COUNTYFP, 3))

bmi_tract <- obs_sub %>%
  mutate(BMI = as.numeric(VALUE)) %>%
  group_by(TRACTCE) %>%
  summarize(mean_bmi = mean(BMI)) %>%
  mutate(TRACTCE = str_sub(TRACTCE, 6)) %>%
  distinct(TRACTCE, .keep_all = TRUE)
```

Maps
```{r}
create_map("co", bmi, "mean_bmi")
create_map("tr", bmi_tract, "mean_bmi")
```


Food Desert Data
(Look at Erin's)

```{r}
setwd("/Users/chanamessinger/Dropbox/Columbia/Practicum/spring2019/My data")
food_desert <- read_csv("FoodDesert.csv")
CountyName <- c("Barnstable", "Berkshire", "Bristol", "Dukes", "Essex", "Franklin", "Hampden", "Hampshire", "Middlesex", "Nantucket", "Norfolk", "Plymouth", "Suffolk", "Worcester")
COUNTYFP <- c("001", "003", "005", "007", "009", "011", "013", "015", "017", "019", "021", "023", "025", "027")
mass_counties <- as.data.frame(cbind(CountyName,COUNTYFP))
food_desert_sub <- food_desert %>%
  filter(State == "MA")

food_desert_sub <- left_join(food_desert_sub, mass_counties, "CountyName")

food_desert_sub <- food_desert_sub %>%
  group_by(COUNTYFP) %>%
  summarize(perlowaccess = sum(LOWA_POP)/sum(TOTALPOP))

bmi_food_data <- left_join(bmi, food_desert_sub)
```

```{r}
create_map("co", bmi_food_data, "perlowaccess")
create_map("co", bmi_food_data, "mean_bmi")
```
```{r}
setwd("/Users/chanamessinger/Dropbox/Columbia/Practicum/spring2019/My data/Shape Files")
mass_census <- readOGR("mass_census/.","cb_2017_25_tract_500k")
  View(mass_census@data)
  
shape_with_data <- mass_census@data %>% 
  left_join(bmi_food_data, by = "COUNTYFP")
  
mass_census@data <- shape_with_data

merged <- geo_join(bmi_food_data, mass_census, "COUNTYFP")

pal <- colorNumeric("Reds", domain=bmi_food_data$mean_bmi)
pal2 <- colorNumeric("Blues", domain=bmi_food_data$perlowaccess)


mass_census@data %>% 
  leaflet() %>% 
  addTiles() %>% 
  addPolygons(data = mass_census,
              fillColor = ~pal(mass_census@data$mean_bmi),
              fillOpacity = 1,
              popup = ~shape_with_data$perlowaccess) %>%
  addLegend(pal = pal, 
            values = mass_census$mean_bmi, 
            position = "bottomright", 
            title = "Mean BMI In Each County")

# mass_census@data %>% 
#   leaflet() %>% 
#   addTiles() %>% 
#   addMarkers(data = select(mass_census@data, lat, long, COUNTYFP, mean_bmi),
#               group = "bmi") %>%
#   addMarkers(data = select(mass_census, COUNTYFP, perlowaccess),
#               fillColor = ~pal2(mass_census@data$perlowaccess),
#               fillOpacity = 1,
#               group = "food") %>%
#   addLegend(pal = pal, 
#             values = mass_census$mean_bmi, 
#             position = "bottomright", 
#             title = "Mean BMI In Each County")
```

