---
title: "PART I. Identifying BC & pregnancy"
author: "Angie Zhang (incorporated Erin's data exploration)"
date: "March 5th 2019"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Step #1. Filter patients by gender 
```{r}
# Load in 'patients' data
patients<-read.csv(file.choose()) 
```

```{r}
library(dplyr)
library(tidyr)

# Identify female patients
female <- patients %>%
  filter(GENDER=="F")

count(female) 
head(female)
```
There are a total of 721 female patients from our observatons


Step #2. Identify BCs vs.non-BC - confirm on gender

The codes assigned to birth control are as follows (sorted by type of bc): 
(1) Injections:
  * 1000158 = 0.65 ML depo-subQ provera 160 MG/ML Prefilled Syringe
  * 1000128 = 1 ML Depo-Provera 150 MG/ML Injection
(2) Pills:
  * 748962 = Camila 28 Day Pack
  * 831533 = Errin 28 Day Pack
  * 749785 = Ortho Tri-Cyclen 28 Day Pack
  * 749762 = Seasonique 91 Day Pack
  * 751905 = Trinessa 28 Day Pack
  * 748856 = Yaz 28 Day Pack* 1359133 = Estrostep Fe 28 Day Pack
  * 757594 = Jolivette 28 Day Pack
  * 748879 = Levora 0.15/30 28 Day Pack
  * 235389 = Mestranol / Norethynodrel [Enovid]
(3) IUD/implants:
  * 1605257 = Liletta 52 MG Intrauterine System
  * 1856546 = Kyleena 19.5 MG Intrauterine System
  * 807283 = Mirena 52 MG Intrauterine System
  * 1111011 = Nexplanon 68 MG Drug Implant
  * 646250 = Implanon 68 MG Drug Implant
(4) Viginal Ring:
  * 1367439 = NuvaRing 0.12/0.015 MG per 24HR 21 Day Vaginal Ring

```{r}
# Load in 'medications' data
meds<-read.csv(file.choose()) 
```

```{r, eval=TRUE, echo=FALSE}
# label birth control using codes identified above
bc <- c(1000158, 1000128, 748962, 831533, 1359133, 646250, 757594, 1856546, 748879, 1605257, 235389, 807283, 1111011, 1367439, 749785, 749762, 751905, 748856)
head(meds)

# create a dataframe for those on bc
meds_bc<- meds %>% filter(CODE %in% bc)

# Find out the numer of patients who are/were on bc
meds %>% filter(CODE %in% bc) %>% summarise(n_distinct(PATIENT))

# check to ensure all those who use BC are actually females 
meds_bc$PATIENT[!(meds_bc$PATIENT %in% female$ID)]

```
There are 246 unique patients who took birth control, whom are all confirmed to be females. 


Step #3. Create indicator variable for birth control: 1 if on BC, 0 if not.

```{r}
meds_bc$bc_yes <- rep(1, nrow(meds_bc))
patient_bc <- merge(x=female, y=meds_bc[,c("PATIENT","bc_yes")], by.x="ID", by.y="PATIENT", all.x = TRUE)

patient_bc$bc_yes <- transform(patient_bc, bc_yes=as.numeric(bc_yes))

patient_bc$bc_yes <- ifelse(patient_bc$bc_yes==1,1,0)

names(patient_bc)
```


Step #4, Examine characteristics of BC paitents.
   (1) Overall: characterstics of those who use BC (by race, age, marital status)
   (2) By category: pills, IUD, Viginal ring, Inplant, injection,

```{r}
# add race and marital status columns on meds_bc file, showing the demographic info of those on bc
med_demo <- merge(x=meds_bc, y=patients[,c("ID","RACE","MARITAL")], by.x="PATIENT", by.y="ID", all.x = TRUE)
head(med_demo)
med_race<- med_demo%>%
  group_by(RACE)%>%
  tally
med_race
barplot(med_race$n,main = "Race Distribution for bc users",
xlab = "Race",
ylab = "# of Patients",
names.arg = c("asian","black","hispanic","white"))
```

```{r}
#marital status

med_marital<- med_demo%>%
  group_by(MARITAL)%>%
  tally
med_marital
barplot(med_marital$n, main="Distribution of marital status for bc users", xlab ="marital status", ylab="# of patients", names.arg=c("N/A","Married","Single"))


```

```{r}
# Calculate patients age
# install.packages('eeptools')
library(eeptools)
library(lubridate)

patients$BIRTHDATE<-ymd(patients$BIRTHDATE, format="%m/%d/%Y")
patients$DEATHDATE<-ymd(patients$DEATHDATE, format="%m/%d/%Y")
str(patients$BIRTHDATE)
patients%>% 
  mutate(age= as.period(interval(patients$BIRTHDATE, patients$DEATHDATE), unit = "year"))
  as.numeric(age)
head(age)

#mutate( age = age_calc(patients$BIRTHDATE, enddate = Sys.Date(), units='years', precise = FALSE)
#patients$age <- age_calc(dob=patients$BIRTHDATE, enddate = patients$DEATHDATE, units='years', precise = FALSE)
```

  
  
Step #5. Identify Pregnant vs.non-Pregnant (using Careplan, Conditions and Encounter Dataset )

### Careplan
```{r}
# load Careplan file
care<-read.csv(file.choose()) 
```

```{r, eval=TRUE, echo=FALSE}
# Looking at normal pregnancies within Careplans dataset.
library(dplyr)
normpreg <- care %>%
  filter(REASONDESCRIPTION=="Normal pregnancy")
normpreg %>% summarize(n_distinct(PATIENT))
```
In the Careplans csv, there are 167 pregnancies. The condition descriptions for patients without any "ReasonDescription," by and large, do not appear to reflect pregnancies. 

### Conditions
```{r}
# Load in conditions data
cond<-read.csv(file.choose()) 
```

An initial review of conditions data revealed there are 10 descriptions that could relate to pregenancy.
```{r, eval=TRUE, echo=FALSE}
obs <- c("Antepartum eclampsia", "Blighted ovum", "Complication occuring during pregnancy", "Congenital uterine anomaly", "Fetus with chromosomal abnormality", "Miscarriage in first trimester", "Miscarriage in second trimester", "Normal pregnancy", "Preeclampsia", "Tubal pregnancy")
allpreg <- cond %>%
  filter(DESCRIPTION %in% obs)
allpreg %>% summarize(n_distinct(PATIENT))
```
In the Conditions csv, there are 139 unique patients with pregnancy-related descriptions. 

### Encounters
```{r}
# Load in encounters data
encount<-read.csv(file.choose()) 
```

In the Encounters csv, three Descriptions seem specific to pregnancies, "Obstretric emergency hospital admission," "Prenatal initial visit," and "Prenatal visit." When applying the same pregnancy-related categories used for the Conditions csv to the ReasonDescription column, "Admission to surgical department" and "Patient-initiated encounter" were highlighted, indicating that the three categories I initially identified do not comprehensively select all pregnancy-related encounters. The following code tries to account for that. 

```{r, eval=TRUE, echo=FALSE}
obs2 <- c("Obstretric emergency hospital admission", "Prenatal initial visit", "Prenatal visit")
pregencount <- encount %>%
  filter(DESCRIPTION %in% obs2 | REASONDESCRIPTION %in% obs)
pregencount %>% summarise(n_distinct(PATIENT))
```
Per the Encounters csv, there are 135 patients who attended pregnancy-related doctor's appointments (which is taken as a proxy for pregnancy).

## Merging datasets
combine the three datasets from above to find out the number of unique pregnancies.
```{r, eval=TRUE, echo=FALSE}
half <- merge(x=pregencount, y=normpreg, by="PATIENT", all = TRUE) #combine encounter with careplan data

colnames(half)[2]<-"enc.ID"
colnames(half)[3]<-"enc.DATE"
colnames(half)[4]<-"enc.CODE"
colnames(half)[5] <- "enc.DESCRIPTION"
colnames(half)[6] <-"enc.REASONCODE"
colnames(half)[7] <-"enc.REASONDESCRIPTION"
colnames(half)[8] <-"cp.ID"
colnames(half)[9] <-"cp.START"
colnames(half)[10] <-"cp.STOP"
colnames(half)[11] <-"cp.ENCOUNTER"
colnames(half)[12] <-"cp.CODE"
colnames(half)[13] <-"cp.DESCRIPTION"
colnames(half)[14]<-"cp.REASONCODE"
colnames(half)[15] <-"cp.REASONDESCRIPTION"

names(half)

all <- merge(x=half, y=allpreg, by="PATIENT", all = TRUE) #combine with conditions data

colnames(all)[16]<-"cond.START"
colnames(all)[17]<-"cond.STOP"
colnames(all)[18]<-"cond.ENCOUNTER"
colnames(all)[19]<-"cond.CODE"
colnames(all)[20]<-"cond.DESCRIPTION"
names(all)
all %>% summarize(n_distinct(PATIENT))
```
There are 191 unique pregnancies amongst the three spreadsheets. 


Step #6. BC to Pregnant
```{r, eval=TRUE, echo=FALSE}
meds_bc <- meds %>%
  filter(CODE %in% bc)
names(meds_bc)[names(meds_bc) == "CODE"] <- "CODE.bc"
names(meds_bc)[names(meds_bc) == "DESCRIPTION"] <- "DESCRIPTION.bc"
both <- merge(x=meds_bc, y=all, by="PATIENT")
both %>% summarize(n_distinct(PATIENT))
```
141 women who are on birth control got pregnant (or at least had a pre-pregnancy consultation).


Step #7. Check for multiple instances of pregnancy 
```{r}
library(psych)
names(all)
all$cp.START <- as.Date.POSIXct(all$cp.START, tryFormats = "%m/%d/%Y")
all$cp.STOP <- as.Date.POSIXct(all$cp.STOP, tryFormats = "%m/%d/%Y")
all$cond.START <- as.Date.POSIXct(all$cond.START, tryFormats = "%m/%d/%Y")
all$cond.STOP <- as.Date.POSIXct(all$cond.STOP, tryFormats = "%m/%d/%Y")
str(all)

all%>%
  group_by(PATIENT) %>%
  with(., difftime(max(cp.START), min(cp.STOP)))



all$df_cp <- all$cp.STOP - all$cp.START
all$df_cond <- all$cond.STOP - all$cond.START

describe(all$df_cp)

all%>%
  group_by(PATIENT) %>%
  summarize(cp_days=sum(df_cp),cond_days=sum(df_cond))

head(all$df_cp)
head(all$df_cond)

```

Min and max dates that indicate that (>9months)







